---
title: "predicting mortality"
author: "Thomas Benacci"
date: "2025-11-20"
output: html_document
---

# Libraries

```{r, include=FALSE}
library(tidyverse)
library(survival)
library(ranger)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(data.table)
library(glmnet) # lasso Cox regression
library(broom) # tidy extraction
```

# Data

```{r}
data <- fread(
  "Survey_1_feature_data.csv",
  data.table = F
)
# Identify predictor columns (all char_ variables except demographics used as adjustments)
demo_vars <- c("char_Age", "char_Male", "char_Black", "char_ForeignBorn")
all_char_vars <- grep("^char_", names(data), value = TRUE)
predictor_vars <- setdiff(all_char_vars, demo_vars)
cat("Number of predictors:", length(predictor_vars), "\n")
cat("Demographics:", paste(demo_vars, collapse = ", "), "\n")
```

# Kaplan-Meier Survival Curve

```{r}
km_fit <- survfit(Surv(Max_Age, DECEASED) ~ 1, data = data)

autoplot(km_fit) +
  labs(
    title = "Kaplan–Meier Survival Curve",
    x = "Age",
    y = "Survival Probability"
  ) +
  theme_minimal()
```

# Independent Cox Proportional Hazards Models (Replicating Paper Fig. 1 / Table 1)

Each predictor is tested independently, adjusted for demographics only. Cluster-robust SEs by household (M2FAMNUM). Bonferroni-corrected CIs.

```{r}
n_predictors <- length(predictor_vars)
bonferroni_alpha <- 0.05 / n_predictors
z_bonf <- qnorm(1 - bonferroni_alpha / 2)
# Run independent Cox models — one per predictor, adjusted for demographics
independent_results <- lapply(predictor_vars, function(pred) {
  # Build formula: Surv(Max_Age, DECEASED) ~ predictor + demographics + cluster(M2FAMNUM)
  fml <- as.formula(paste0(
    "Surv(Max_Age, DECEASED) ~ ", pred, " + ",
    paste(demo_vars, collapse = " + "),
    " + cluster(M2FAMNUM)"
  ))
  fit <- tryCatch(
    coxph(fml, data = data),
    error = function(e) NULL
  )
  if (is.null(fit)) return(NULL)
  s <- summary(fit)
  # Extract the row for our predictor of interest (not the demographics)
  coef_row <- s$coefficients[pred, , drop = FALSE]
  hr <- exp(coef_row[, "coef"])
  se <- coef_row[, "robust se"]
  log_hr <- coef_row[, "coef"]
  p_val <- coef_row[, "Pr(>|z|)"]
  # Standard 95% CI
  ci_lo_95 <- exp(log_hr - 1.96 * se)
  ci_hi_95 <- exp(log_hr + 1.96 * se)
  # Bonferroni-corrected CI
  ci_lo_bonf <- exp(log_hr - z_bonf * se)
  ci_hi_bonf <- exp(log_hr + z_bonf * se)
  
  data.frame(
    Variable = pred,
    HR = hr,
    CI_lo_95 = ci_lo_95,
    CI_hi_95 = ci_hi_95,
    CI_lo_bonf = ci_lo_bonf,
    CI_hi_bonf = ci_hi_bonf,
    p_value = p_val,
    p_bonf_sig = p_val < bonferroni_alpha,
    stringsAsFactors = FALSE
  )
})
# Combine and sort by HR descending
independent_df <- bind_rows(independent_results) %>%
  arrange(desc(HR))
# Clean up variable names for display
independent_df$Variable_clean <- independent_df$Variable %>%
  str_remove("^char_") %>%
  str_replace_all("([a-z])([A-Z])", "\\1 \\2") %>%
  str_replace_all("_", " ")
# Print ranked results
independent_df %>%
  select(Variable_clean, HR, CI_lo_95, CI_hi_95, CI_lo_bonf, CI_hi_bonf, p_value, p_bonf_sig) %>%
  print(n = 50)
```

# Forest Plot of Independent Cox Regression HRs (Fig. 1 Replication)

```{r, fig.width=10, fig.height=12}
# Assign domains for color coding (matching paper's 6 domains)
domain_map <- tribble(
  ~Variable, ~Domain,
  "char_LowFatherOccStat",         "Childhood Adversity",
  "char_LowMotherOccStat",         "Childhood Adversity",
  "char_LowFatherEdu",             "Childhood Adversity",
  "char_LowMotherEdu",             "Childhood Adversity",
  "char_LowerIncome",              "Adulthood Socioeconomic",
  "char_LowerWealth",              "Adulthood Socioeconomic",
  "char_LowerEdu",                 "Adulthood Socioeconomic",
  "char_LowerOccStat",             "Adulthood Socioeconomic",
  "char_RecentFinDiff",            "Adulthood Socioeconomic",
  "char_HistOfUnemployment",       "Adulthood Socioeconomic",
  "char_HistOfRenting",            "Adulthood Socioeconomic",
  "char_HistOfMedicaid",           "Adulthood Socioeconomic",
  "char_LowNeighborCohesion",      "Adulthood Socioeconomic",
  "char_AlcoholAbuse",             "Adulthood Health Behaviors",
  "char_CurrentSmoker",            "Adulthood Health Behaviors",
  "char_HistOfSmoking",            "Adulthood Health Behaviors",
  "char_LowNoModActivity",         "Adulthood Health Behaviors",
  "char_LowNoVigActivity",         "Adulthood Health Behaviors",
  "char_SleepProblems",            "Adulthood Health Behaviors",
  "char_HistOfDivorce",            "Adulthood Social Connections",
  "char_NeverMarried",             "Adulthood Social Connections",
  "char_LowerPosIntWithFamily",    "Adulthood Social Connections",
  "char_NegIntWithFamily",         "Adulthood Social Connections",
  "char_LowerPosIntWithFriends",   "Adulthood Social Connections",
  "char_NegIntWithFriends",        "Adulthood Social Connections",
  "char_LowerAgreeableness",       "Adulthood Psychological",
  "char_LowerExtroversion",        "Adulthood Psychological",
  "char_LowerNeuroticism",         "Adulthood Psychological",
  "char_LowerConscientiousness",   "Adulthood Psychological",
  "char_LowerOpennessToExperiences","Adulthood Psychological",
  "char_Hopelessness",             "Adulthood Psychological",
  "char_LowerLifeSatisfaction",    "Adulthood Psychological",
  "char_LowerOptimism",            "Adulthood Psychological",
  "char_LowerPosAffectivity",      "Adulthood Psychological",
  "char_NegAffectivity",           "Adulthood Psychological",
  "char_LowerPurposeInLife",       "Adulthood Psychological",
  "char_LowerReligiosity",         "Adulthood Psychological",
  "char_LowerSenseOfMastery",      "Adulthood Psychological",
  "char_DailyDiscrimination",      "Adulthood Adverse Experiences",
  "char_MajorDiscrimination",      "Adulthood Adverse Experiences"
)
#
plot_df <- independent_df %>%
  left_join(domain_map, by = "Variable") %>%
  mutate(
    Variable_clean = factor(Variable_clean, levels = rev(Variable_clean)) # ordered by HR
  )
#
ggplot(plot_df, aes(x = HR, y = Variable_clean, color = Domain)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = CI_lo_bonf, xmax = CI_hi_bonf), height = 0.3) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Independent Cox Regression Hazard Ratios (MIDUS)",
    subtitle = paste0("Bonferroni-corrected 95% CIs (", n_predictors, " comparisons). Age as baseline hazard."),
    x = "Hazard Ratio",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 8)
  )
```

# Variance Explained by Domain

```{r}
# Pseudo R-squared (Royston & Sauerbrei) for each domain
domains_list <- split(domain_map$Variable, domain_map$Domain)
domain_r2 <- lapply(names(domains_list), function(d) {
  vars_in_domain <- domains_list[[d]]
  # Ensure all vars exist in data
  vars_in_domain <- vars_in_domain[vars_in_domain %in% names(data)]
  fml <- as.formula(paste0(
    "Surv(Max_Age, DECEASED) ~ ",
    paste(c(vars_in_domain, demo_vars), collapse = " + ")
  ))
  fit <- coxph(fml, data = data)
  s <- summary(fit)
  data.frame(
    Domain = d,
    R2_Royston = s$rsq[1],  # generalized R^2
    stringsAsFactors = FALSE
  )
})
# Demographics-only baseline
demo_fit <- coxph(
  as.formula(paste0("Surv(Max_Age, DECEASED) ~ ", paste(demo_vars, collapse = " + "))),
  data = data
)
demo_r2 <- summary(demo_fit)$rsq[1]
domain_r2_df <- bind_rows(domain_r2) %>%
  mutate(R2_above_demo = R2_Royston - demo_r2) %>%
  arrange(desc(R2_above_demo))
cat("\nDemographics-only R²:", round(demo_r2, 4), "\n\n")
print(domain_r2_df)
# All predictors combined
all_fml <- as.formula(paste0(
  "Surv(Max_Age, DECEASED) ~ ",
  paste(c(predictor_vars, demo_vars), collapse = " + ")
))
all_fit <- coxph(all_fml, data = data)
cat("\nAll predictors combined R²:", round(summary(all_fit)$rsq[1], 4), "\n")
cat("Incremental above demographics:", round(summary(all_fit)$rsq[1] - demo_r2, 4), "\n")
```

# Lasso Cox Regression (Replicating Paper Fig. 2)

All predictors fit simultaneously with penalized likelihood.

```{r, fig.width=10, fig.height=8}
# Prepare model matrix (all predictors + demographics)
all_model_vars <- c(predictor_vars, demo_vars)
# Remove rows with any NA in predictors OR response before building Surv object
complete_idx <- complete.cases(data[, c("Max_Age", "DECEASED", all_model_vars)])
data_complete <- data[complete_idx, ]
x_mat <- as.matrix(data_complete[, all_model_vars])
y_surv <- Surv(data_complete$Max_Age, data_complete$DECEASED)
cat("Complete cases for lasso:", nrow(data_complete), "of", nrow(data), "\n")
# Cross-validated lasso
set.seed(42)
cv_fit <- cv.glmnet(x_mat, y_surv, family = "cox", alpha = 1, nfolds = 10)
# Extract coefficients at lambda.min
lasso_coefs <- coef(cv_fit, s = "lambda.min")
lasso_df <- data.frame(
  Variable = rownames(lasso_coefs),
  coef = as.numeric(lasso_coefs),
  HR = exp(as.numeric(lasso_coefs)),
  stringsAsFactors = FALSE
) %>%
  filter(coef != 0) %>%
  arrange(desc(HR))
# Clean names for display
lasso_df$Variable_clean <- lasso_df$Variable %>%
  str_remove("^char_") %>%
  str_replace_all("([a-z])([A-Z])", "\\1 \\2") %>%
  str_replace_all("_", " ")
lasso_df <- lasso_df %>%
  left_join(
    domain_map %>% mutate(Variable_clean = Variable %>%
      str_remove("^char_") %>%
      str_replace_all("([a-z])([A-Z])", "\\1 \\2") %>%
      str_replace_all("_", " ")),
    by = "Variable_clean"
  ) %>%
  # Mark demographic vars
  mutate(Domain = ifelse(is.na(Domain), "Sociodemographic", Domain))
# Plot
lasso_df %>%
  mutate(Variable_clean = factor(Variable_clean, levels = rev(Variable_clean))) %>%
  ggplot(aes(x = HR, y = Variable_clean, color = Domain)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Hazard Ratios from Lasso Cox Regression (MIDUS)",
    subtitle = "Shown only for HRs that differ from 1 (non-zero coefficients)",
    x = "Hazard Ratio",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 9)
  )
# Print lasso results table
lasso_df %>%
  select(Variable_clean, HR, coef, Domain) %>%
  print(n = 50)
```

# Random Forest Survival Analysis (Replicating Paper SI Fig. S1)

```{r, fig.width=10, fig.height=10}
# Prepare data for ranger (no NAs allowed)
rf_data <- data %>%
  select(Max_Age, DECEASED, all_of(all_model_vars)) %>%
  drop_na()
# Fit random survival forest
set.seed(42)
rf_fit <- ranger(
  Surv(Max_Age, DECEASED) ~ .,
  data = rf_data,
  num.trees = 500,
  importance = "permutation",
  mtry = floor(sqrt(length(all_model_vars))),
  min.node.size = 20
)
# Variable importance
vimp <- data.frame(
  Variable = names(rf_fit$variable.importance),
  Importance = rf_fit$variable.importance,
  stringsAsFactors = FALSE
) %>%
  arrange(desc(Importance))
#
vimp$Variable_clean <- vimp$Variable %>%
  str_remove("^char_") %>%
  str_replace_all("([a-z])([A-Z])", "\\1 \\2") %>%
  str_replace_all("_", " ")
#
vimp <- vimp %>%
  left_join(
    domain_map %>% mutate(Variable_clean = Variable %>%
      str_remove("^char_") %>%
      str_replace_all("([a-z])([A-Z])", "\\1 \\2") %>%
      str_replace_all("_", " ")),
    by = "Variable_clean"
  ) %>%
  mutate(Domain = ifelse(is.na(Domain), "Sociodemographic", Domain))
#
vimp %>%
  mutate(Variable_clean = factor(Variable_clean, levels = rev(Variable_clean))) %>%
  ggplot(aes(x = Importance, y = Variable_clean, fill = ifelse(Importance > 0, "Positive", "Negative"))) +
  geom_col() +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "firebrick"), name = "Direction") +
  labs(
    title = "Random Forest Survival Analysis — Variable Importance (MIDUS)",
    subtitle = "Permutation importance. Positive = improves prediction; Negative = likely noise.",
    x = "Variable Importance",
    y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))
```

# Personal Prediction Example

```{r}
# Pick a random LIVING row
set.seed(123)
new_obs <- data %>%
  filter(DECEASED == 0) %>%
  slice_sample(n = 1)
# Refit full multivariate Cox model (no cluster term for survfit prediction)
cox_full <- coxph(
  as.formula(paste0(
    "Surv(Max_Age, DECEASED) ~ ",
    paste(c(predictor_vars, demo_vars), collapse = " + ")
  )),
  data = data,
  model = TRUE,
  x = TRUE
)
# Predict survival curve
surv_curve <- survfit(cox_full, newdata = new_obs)
# Convert S(t) -> probability of dying at age t
times <- surv_curve$time
S     <- surv_curve$surv
death_probs <- S[-length(S)] - S[-1]
death_dist <- data.frame(
  age = times[-length(times)],
  prob_death_at_age = death_probs
)
# Normalize
death_dist$prob_death_at_age <- death_dist$prob_death_at_age /
                                sum(death_dist$prob_death_at_age)
# Cumulative probability of death by age
death_dist <- death_dist %>%
  arrange(age) %>%
  mutate(cum_prob = cumsum(prob_death_at_age))
ggplot(death_dist, aes(x = age, y = cum_prob)) +
  geom_line() +
  labs(
    title = "Cumulative Probability of Death by Age (Example Individual)",
    x = "Age",
    y = "Cumulative Probability"
  ) +
  theme_minimal()
```

# Save Model

```{r}
saveRDS(cox_full, "cox_mortality_model.rds")
saveRDS(list(predictor_vars = predictor_vars, demo_vars = demo_vars), "model_variable_names.rds")
cat("Model and variable names saved.\n")
```
